A Higher-Order Approach to Explicit Contexts in LF

### 2015-03-08 paper/paper.tex
tweaks in admissibility proof formatting
typo fix
s/proving/justifying
typos

To encode this in a straightforward, higher-order manner, we shall first
stipulate that each variable in the telescope will be an LF-variable. Then,
each type $A_i$ in the telescope will be a term with $i-n$ LF-bindings (i.e.\ it
will be an $i$-ary $\lambda$-abstraction). We'll define by mutual induction a
sort for telescopes $\sorttele$, and a sort for open terms with respect to a
telescope $\sortoexp{\Psi}$:

@@

The disjoint union of a family of sets is also easily encoded; first
Martin-L\"of's rules:
\begin{gather*}
  \infer{
    \isset{(\mathsf{\Sigma}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }\quad
  \infer{
    \mem{(a,b)}{(\mathsf{\Sigma}x\in A)B}
  }{
    \mem{a}{A} &
    \mem{b}{[a/x]B}
  }\quad
  \infer{
    \eqmem{(a,b)}{(c,d)}{(\mathsf{\Sigma}x\in A)B}
  }{
    \eqmem{a}{c}{a} &
    \eqmem{b}{d}{[a/x]B}
  }
\end{gather*}
%
and their encoding:
\begin{gather*}
  \infer{
    \canset{\tysg(A;B)}
  }{
    \isset{A} &
    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
  }\quad
  \infer{
    \canmem{\pair(M,N)}{\tysg(A;B)}
  }{
    \mem{M}{A} &
    \mem{N}{B(M)}
  }\quad
  \infer{
    \eqcanmem{\pair(M;N)}{\pair(M';N')}{\typi(A;B)}
  }{
    \eqmem{M}{M'}{A} &
    \eqmem{N}{N'}{B(M)}
  }
\end{gather*}

And what of the elimination rules? The entire point of the verificationist
meaning explanations which pervade \cite{Martin-Lof-1979} is that the
elimination rules will not be part of the definitions of the types, but rather
rules which are admissible, justified by the meanings of the judgements and the
definitions of the types. So, only after having propounded the definitions
above may we entertain \emph{proving} the elimination rules in the logical
framework.

@@

  Admissibility follows from $\mathcal{D}'$, $\mathcal{D}''$, $\mathcal{E}$;
  the derivation $\mathcal{D''}(N)(\mathcal{E})$ must have a canonical form as follows:
  \[
    \infer{
      \mem{E(N)}{B(N)}
    }{
      \begin{array}{c}
        \mathcal{F}\\
        \eval{E(N)}{X}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}'\\
        \eval{B(N)}{C}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}''\\
        \canmem{X}{C}
      \end{array}
    }
  \]
  Whence we may construct the goal:
  \[
    \infer{
      \mem{\ap(M;N)}{B(N)}
    }{
      \infer{
        \eval{\ap(M;N)}{X}
      }{
        \infer[\mathcal{D}']{\eval{M}{\lam E}}{} &
        \infer[\mathcal{F}]{\eval{E(N)}{X}}{}
      }&
      \infer[\mathcal{F}']{\eval{B(N)}{C}}{}&
      \infer[\mathcal{F}'']{\canmem{X}{C}}{}&
    }
  \]
\end{proof}

\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
\[
  \infer[\tysg_{E_1}]{
    \mem{\fst(M)}{A}
  }{
    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
  }\quad
  \infer[\tysg_{E_2}]{
    \mem{\snd(M)}{B(\fst(M))}
  }{
    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
  }
\]
}
\begin{proof}
\end{proof}

>>>>

To encode this in a straightforward, higher-order manner, we shall first
stipulate that each variable in the telescope will be an LF-variable. Then,
each type $A_i$ in the telescope will be a term with $i-1$ LF-bindings (i.e.\ it
will be an $(i-1)$-ary $\lambda$-abstraction). We'll define by mutual induction a
sort for telescopes $\sorttele$, and a sort for open terms with respect to a
telescope $\sortoexp{\Psi}$:

@@

The disjoint union of a family of sets is also easily encoded; first
Martin-L\"of's rules:
\begin{gather*}
  \infer{
    \isset{(\mathsf{\Sigma}x\in A)B}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }\quad
  \infer{
    \mem{(a,b)}{(\mathsf{\Sigma}x\in A)B}
  }{
    \mem{a}{A} &
    \mem{b}{[a/x]B}
  }\quad
  \infer{
    \eqmem{(a,b)}{(c,d)}{(\mathsf{\Sigma}x\in A)B}
  }{
    \eqmem{a}{c}{A} &
    \eqmem{b}{d}{[a/x]B}
  }
\end{gather*}
%
and their encoding:
\begin{gather*}
  \infer{
    \canset{\tysg(A;B)}
  }{
    \isset{A} &
    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
  }\quad
  \infer{
    \canmem{\pair(M,N)}{\tysg(A;B)}
  }{
    \mem{M}{A} &
    \mem{N}{B(M)}
  }\quad
  \infer{
    \eqcanmem{\pair(M;N)}{\pair(M';N')}{\typi(A;B)}
  }{
    \eqmem{M}{M'}{A} &
    \eqmem{N}{N'}{B(M)}
  }
\end{gather*}

And what of the elimination rules? The entire point of the verificationist
meaning explanations which pervade \cite{Martin-Lof-1979} is that the
elimination rules will not be part of the definitions of the types, but rather
rules which are admissible, justified by the meanings of the judgements and the
definitions of the types. So, only after having propounded the definitions
above may we entertain \emph{justifying} the elimination rules in the logical
framework.

@@

  Admissibility follows from $\mathcal{D}'$, $\mathcal{D}''$, $\mathcal{E}$;
  the derivation $\mathcal{D''}(N)(\mathcal{E})$ must have a canonical form as follows:
  \[
    \infer[\mathcal{D''}(N)(\mathcal{E})]{
      \mem{E(N)}{B(N)}
    }{
      \begin{array}{c}
        \mathcal{F}\\
        \eval{E(N)}{X}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}'\\
        \eval{B(N)}{C}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}''\\
        \canmem{X}{C}
      \end{array}
    }
  \]
  Whence we may construct the goal:
  \[
    \infer{
      \mem{\ap(M;N)}{B(N)}
    }{
      \infer{
        \eval{\ap(M;N)}{X}
      }{
        \begin{array}{c}
          \mathcal{D}'\\
          \eval{M}{\lam E}
        \end{array} &
        \begin{array}{c}
          \mathcal{F}\\
          \eval{E(N)}{X}
        \end{array}
      }&
      \begin{array}{c}
        \mathcal{F}'\\
        \eval{B(N)}{C}
      \end{array}&
      \begin{array}{c}
        \mathcal{F}''\\
        \canmem{X}{C}
      \end{array}
    }
  \]
\end{proof}

\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
\[
  \infer[\tysg_{E_1}]{
    \mem{\fst(M)}{A}
  }{
    \mem{M}{\thunk{\tysg(A;B)}} &
  }\quad
  \infer[\tysg_{E_2}]{
    \mem{\snd(M)}{B(\fst(M))}
  }{
    \mem{M}{\thunk{\tysg(A;B)}} &
  }
\]
}
\begin{proof}
\end{proof}
